Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-18 bbbq_article.tex
Add constants
Move information from figures to table

[GVDB: A narrative was missing a bit. Why was this done?]
To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
we reanalyzed published peptide elution studies.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in MHC-I and MHC-II elution 
studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} respectively), 
for the two topology prediction tools TMHMM and 
PureseqTM (\cite{krogh2001predicting} and \cite{wang2019efficient} respectively). 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
[GVDB: Also add a conclusion at the end of the paragraph.]
From these findings, we conclude that PureseqTM and TMHMM 
result in prediction of a similar number of TMHs 
and that peptides derived from TMHs have been identified for both MHC-I and -II.

@@

Where TMH-derived epitopes are over-presented by MHC-I,
we wondered if this finding would also generalize to MHC-II.
Figure \ref{fig:2} shows the percentages of MHC-II epitopes overlapping 
with TMHs for our human, viral and bacterial proteome.
See the supplementary materials (table \ref{tab:tmh_binders_mhc2}) 
for the TMH and epitope counts.
We show that MHC-II has the same pattern of over-presentation
as MHC-I, yet this over-presentation is more extreme.
Also for MHC-II, we observe that the extent the which TMH-derived
epitopes are presented is similar between the haplotypes, regardless
of the protein source.

@@

In the final set of experiments, 
we addressed the question whether TMHs 
would be more evolutionary conserved than soluble proteins regions 
by comparing the occurrences of SNPs in the genome coding 
for TMHs and soluble protein regions within membrane proteins. 
We obtained 1,129 gene names associated with the phrase 'membrane protein',
which are genes for both membrane-associated proteins (MAPs, no TMH) and 
transmembrane proteins (TMPs, at least 1 TMH).
These genes are linked to 4,811 protein isoforms, 
of which 2,568 are predicted to be TMPs (hence 
2,243 proteins are predicted to be MAPs).
We obtained 38,882 SNPs that resulted in an
amino acids substitution, of which 21,738 were located 
in predicted TMPs.
Per protein, we calculated two percentages: 
(1) the percentage of the protein coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without MAPs,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs and soluble protein domains.
Our analysis shows that SNPs occur more often in soluble domains
than in TMHs.
The effect size is that per 1 SNP found in soluble protein
domains, one finds 0.925 SNPs in TMHs.

@@

@@

@@

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_f_tmh_ncbi.png}
  \caption{
    Distribution of the percentages of TMH of transmembrane proteins.
  }
  \label{fig:f_tmh_ncbi}
\end{figure}

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_n_snps_per_tmp.png}
  \caption{
    Histogram of the number of SNPs per trans-membrane protein.
  }
  \label{fig:n_tmhs_per_protein}
\end{figure}

>>>>

To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
we reanalyzed published peptide elution studies.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in MHC-I and MHC-II elution 
studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} respectively),
for the two topology prediction tools TMHMM and 
PureseqTM (\cite{krogh2001predicting} and \cite{wang2019efficient} respectively). 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented by B cells [For Frans: please check
that the article by Bergseng, 2015, 'Different binding motifs of the celiac 
disease-associated HLA molecules DQ2. 5, DQ2. 2, and DQ7. 5 revealed by 
relative quantitative proteomics of endogenous peptide repertoires'
is appropriate. As far as I can see, it is, but I may overlook something. Richel].

@@

Where TMH-derived epitopes are over-presented by MHC-I,
we wondered if this finding would also generalize to MHC-II.
Figure \ref{fig:2} shows the percentages of MHC-II epitopes overlapping 
with TMHs for our human, viral and bacterial proteome.
See the supplementary materials (table \ref{tab:tmh_binders_mhc2}) 
for the TMH and epitope counts.
We show that MHC-II has the same pattern of over-presentation
as MHC-I, yet this over-presentation is more extreme.
Also for MHC-II, we observe that the extent the which TMH-derived
epitopes are presented is similar between the haplotypes, for
a human, bacterial and viral proteome.

@@

In the final set of experiments, 
we addressed the question whether TMHs 
would be more evolutionary conserved than soluble proteins regions 
by comparing the occurrences of SNPs in the genome coding 
for TMHs and soluble protein regions within membrane proteins.
We obtained 1,129 [For Frans: I would enjoy to write 'approx 1k' here, would that
be acceptable? I do refer to a table with the exact values
later on anyways. Richel] gene names associated with the phrase 'membrane protein',
which are genes for both membrane-associated proteins (MAPs, no TMH) and 
transmembrane proteins (TMPs, at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs (hence 
2,237 proteins are predicted to be MAPs).
We obtained 37,630 SNPs that resulted in an
amino acids substitution, of which 21,024 were located 
in predicted TMPs.
See table \ref{tab:ncbi_counts} for an overview of all amounts.

Per protein, we calculated two percentages: 
(1) the percentage of the protein coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without MAPs,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs 
and soluble protein domains.
This analysis gives the first hint that SNPs 
occur more often in soluble domains than in TMHs.

@@

The effect size is that, 
when randomly selecting SNPs, 
per 100 SNPs found in soluble protein domains, 
one finds 91 SNPs in TMHs,
as depicted in figure \ref{fig:conservation}.

@@

We conclude that,
when randomly selecting SNPs, 
for 100 SNPs found in soluble regions of multi-spanners,
one finds 89.6 SNPs in TMHs, as shown by 
figure \ref{fig:conservation_per_spanner}.

@@

See table \ref{tab:ncbi_counts} for an overview of all amounts.

% Label: tab:ncbi_counts
\input{ncbi_peregrine_results/ncbi_counts.tex}

Figure \ref{fig:snps_per_gene_name_ncbi} shows the distribution of the
number of SNPs per gene name, at 2020-12-14.

@@

Figure \ref{fig:f_tmh_ncbi} shows the distribution 
of the percentages of TMH of transmembrane proteins.

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_f_tmh_ncbi.png}
  \caption{
    Distribution of the percentages of TMH of transmembrane proteins.
    Dashed vertical line: mean \% TMH in all proteins.
    Dotted vertical line: mean \% TMH in TMPs.
  }
  \label{fig:f_tmh_ncbi}
\end{figure}

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_n_snps_per_tmp.png}
  \caption{
    Histogram of the number of SNPs per trans-membrane protein.
    Dashed vertical line: average number of SNPs per TMP
  }
  \label{fig:n_tmhs_per_protein}
\end{figure}
